---
title: "Income Data Description"
output: html_document
---

# Income Data Processing

This script processes income data (ACSST1Y - ACS Selected Economic Characteristics) for each year from 2014-2024, merges all years into a single dataframe where each row represents a year's worth of data, and renames columns using metadata labels.



```{r}
library(tidyverse)
library(dplyr)

## Load a specific year to understand the structure
data_2014 <- read.csv("incomedata/ACSST1Y2014.S1901-Data.csv")
metadata_2014 <- read.csv("incomedata/ACSST1Y2014.S1901-Column-Metadata.csv")

# Display the data and metadata structure
head(data_2014)
head(metadata_2014)
```

## Understanding the Data Structure

In the metadata: each row represents what each column in the actual data refers to. Column codes (like S1901_C01_001E) are matched with their descriptive labels from the metadata.

finally found issue - Each year's data has columns ordered differently. The solution is to match column codes using the metadata, then rename using the descriptive labels. 

```{r}
# Examine the structure more closely
nrow(data_2014)
nrow(metadata_2014)
names(metadata_2014)
```

## Set Up Data Processing

Define the folder path, years available, and read the 2014 metadata to establish standard column ordering.

```{r}
# Define folder path
folder <- "incomedata"

# Define years to process (note: 2020 is missing in the data)
years <- c(2014:2019, 2021:2024)

# reading and using 2014 metadata to label final df columns 
metadata_2014 <- read.csv(paste0(folder, "/ACSST1Y2014.S1901-Column-Metadata.csv"))

# Use the column codes from 2014 as the standard order
# These are codes like S1901_C01_001E, S1901_C01_001M, etc.
standard_names <- metadata_2014$Column.Name

# Display the standard column structure
head(standard_names)
```

## Create Function to Read and Process Each Year

This function reads each year's data, removes the label row (first row), keeps only columns matching 2014's structure, and adds a Year column for tracking temporal changes.

```{r}
# Function to read each year's data and standardize its structure
read_one_year <- function(year, folder, standard_names) {
  
  # Construct file path
  data_file <- paste0(folder, "/ACSST1Y", year, ".S1901-Data.csv")
  
  # Read the data
  data <- read.csv(data_file)
  
  # Remove first row (it contains labels, not actual data values)
  data <- data[-1, ]
  
  # Keep only columns that match 2014 column codes (standardize to 2014 structure)
  # This handles cases where column ordering differs between years
  data <- data[, standard_names]
  
  # Add year column for tracking
  data$Year <- year
  
  return(data)
}

# Test the function on one year
test_data <- read_one_year(2014, folder, standard_names)
nrow(test_data)
head(test_data)
```

## Read All Years

Loop through all available years and store each year's data in a list.

```{r}
# Initialize list to store all years' data
data_list <- list()

# Read each year's data
for (i in 1:length(years)) {
  data_list[[i]] <- read_one_year(years[i], folder, standard_names)
}

# Remove any null entries (in case a year is missing)
data_list <- data_list[!sapply(data_list, is.null)]

# Check how many years were successfully loaded
length(data_list)
```

## Merge All Years into Single Dataframe

Combine all yearly data by binding rows together. Each row now represents one year's worth of data.

```{r}
# Start with the first year
income_all <- data_list[[1]]

# Bind remaining years
for (i in 2:length(data_list)) {
  income_all <- rbind(income_all, data_list[[i]])
}

# Verify the merge
cat("Total rows (years):", nrow(income_all), "\n")
cat("Unique years:", length(unique(income_all$Year)), "\n")
cat("Years included:", sort(unique(income_all$Year)), "\n")
```

## Rename Columns Using Metadata Labels

Replace the column codes (e.g., S1901_C01_001E) with their descriptive labels from the metadata.

```{r}
# Extract labels from metadata
new_names <- metadata_2014$Label

# Rename columns (first n columns are the data columns, last column is Year)
names(income_all)[1:length(new_names)] <- new_names

# Display the renamed structure
head(income_all)
names(income_all)[1:10]  # Show first 10 column names
```

## Data Validation and Summary

```{r}
# Check data consistency
cat("Dataframe dimensions:", nrow(income_all), "rows x", ncol(income_all), "columns\n")
cat("Years in data:", table(income_all$Year), "\n")

# Display column names
cat("Total columns:", ncol(income_all), "\n")
head(names(income_all), 20)

# Display basic data
head(income_all)
```

## Final Processed Dataset

The `income_all` dataframe is now ready for analysis. It contains:
- All years from 2014-2019 and 2021-2024 (2020 not available)
- One row per year of data
- Column names are descriptive labels instead of codes
- A Year column for temporal filtering

```{r}
# View summary of the processed data
summary(income_all[, c(1, 2, ncol(income_all))])
```

