---
title: "Washington Economy Data Description"
output: html_document
---

# Washington Economy Data Processing

This script processes the Washington State economy data (ACSDP1Y - ACS Data Profile) for each year from 2014-2024, merges all years into a single dataframe where each row represents a year's worth of data, and renames columns using metadata labels.


```{r}
library(tidyverse)
library(dplyr)

## Load a specific year to understand the structure
data_2014 <- read.csv("../Wash_Economy_data/ACSDP1Y2014.DP03-Data.csv")
metadata_2014 <- read.csv("../Wash_Economy_data/ACSDP1Y2014.DP03-Column-Metadata.csv")

# Display the data and metadata structure
head(data_2014)
head(metadata_2014)
```

## Understanding the Data Structure

In the metadata: each row represents what each column in the actual data refers to. The column codes (like DP03_0001E) are matched with their descriptive labels.

like the income data, the first row here contains labels (already descriptive), not codes. The actual data starts from the second row.

```{r}
# Examine the structure more closely
nrow(data_2014)
nrow(metadata_2014)
names(metadata_2014)
```

## Set Up Data Processing

Define the folder path, years available, and read the 2014 metadata to establish standard column ordering.

```{r}
# Define folder path
folder <- "../Wash_Economy_data"

# Define years to process (note: 2020 is missing in the data)
years <- c(2014:2019, 2021:2024)

# Read 2014 metadata to find correlating labels for each column
metadata_2014 <- read.csv(paste0(folder, "/ACSDP1Y2014.DP03-Column-Metadata.csv"))

# Use the column codes from 2014 as the standard order

standard_names <- metadata_2014$Column.Name

# Display the standard column structure
head(standard_names)
```

## Create Function to Read and Process Each Year

This function reads each year's data, removes the label row (first row), keeps only columns matching 2014's structure, reorders them to match 2014, and adds a Year column.

```{r}
# Function to read each year's data and standardize its structure
read_one_year <- function(year, folder, standard_names) {
  
  # Construct file path
  data_file <- paste0(folder, "/ACSDP1Y", year, ".DP03-Data.csv")
  
  # Read the data
  data <- read.csv(data_file)
  
  # Remove first row (it contains labels/descriptive names, not actual data values)
  data <- data[-1, ]
  
  # Keep only columns that match 2014 column codes (standardize to 2014 structure)
  # This handles cases where column ordering differs between years
  data <- data[, standard_names]
  
  # Add year column for tracking
  data$Year <- year
  
  return(data)
}

# Test the function on one year
test_data <- read_one_year(2014, folder, standard_names)
nrow(test_data)
head(test_data)
```

## Read All Years

Loop through all available years and store each year's data in a list.

```{r}
# Initialize list to store all years' data
data_list <- list()

# Read each year's data
for (i in 1:length(years)) {
  cat("Processing year", years[i], "\n")
  data_list[[i]] <- read_one_year(years[i], folder, standard_names)
}

# Remove any null entries (in case a year is missing like 2020)
data_list <- data_list[!sapply(data_list, is.null)]


```

## Merge All Years into Single Dataframe

Combine all yearly data by binding rows together. Each row now represents one year's worth of data.

```{r}
# Start with the first year
econ_all <- data_list[[1]]

# Bind remaining years
for (i in 2:length(data_list)) {
  econ_all <- rbind(econ_all, data_list[[i]])
}

# Verify the merge
cat("Total rows (years):", nrow(econ_all), "\n")
cat("Unique years:", length(unique(econ_all$Year)), "\n")
cat("Years included:", sort(unique(econ_all$Year)), "\n")
```

## Rename Columns Using Metadata Labels

Replace the column codes (e.g., DP03_0001E) with their descriptive labels from the metadata.

```{r}
# Extract labels from metadata
new_names <- metadata_2014$Label

# Rename columns (first n columns are the data columns, last column is Year)
names(econ_all)[1:length(new_names)] <- new_names

# Display the renamed structure
head(econ_all)
names(econ_all)[1:10]  # Show first 10 column names
```

## Data Validation and Summary

```{r}
# Check data consistency
cat("Dataframe dimensions:", nrow(econ_all), "rows x", ncol(econ_all), "columns\n")
cat("Years in data:", table(econ_all$Year), "\n")

# Display column names
cat("Total columns:", ncol(econ_all), "\n")
head(names(econ_all), 20)

# Display basic data
head(econ_all)
```

## Final Processed Dataset

The `econ_all` dataframe is now ready for analysis. It contains:
- All years from 2014-2019 and 2021-2024 (2020 not available)
- One row per year of data
- Column names are descriptive labels instead of codes
- A Year column for temporal filtering

```{r}
# View summary of the processed data
summary(econ_all[, c(1, 2, ncol(econ_all))])
```
